Parametricity Consequences of GADTs Representations

### 2021-02-16 20-gadts/GADT-story.tex
added gadt-story-references.bib

There are two standard ways to represent data types when studying
modern functional languages: via their Church encodings in a (possibly
higher-kinded) extension of System F~\cite{gir72}, the calculus at the
core of all such languages, and by augmenting System F
%a Hindley-Milner style calculus
with primitives for constructing them directly as fixpoints.
%{\color{blue} Something about when each of these representations is
%  used. The former perhaps more closely represents how data types are
%  represented in implementations, but the latter are how they have
%  heretofore primarily been treated theoretically.}
In models that validate the program optimization known as {\em short
  cut fusion}~\cite{glp93} the two representations are, in fact,
equivalent, in the sense that they both have the same interpretation
in the model. Models in which the Church encodings and primitive
representations of all {\color{blue} (polynomial)} algebraic data
types (ADTs) are equivalent include the operationally-based model
of~\cite{pit98,pit00} and the categorical models
of~\cite{joh02,joh03}.
%\footnote{{\color{blue} Pitts proves the isomorphism only for
%    polynomial ADTs, whereas Johann {\em et al.}  prove it for any
%    data type that is the fixpoint of a first-order functor.}}
This equivalence was recently extended to the class of what might be
called ``hereditarily polynomial'' {\color{blue} to include truly
  nested such types} nested types via the categorical model
of~\cite{jgj21}. But, as shown in Section~\ref{sec:?} below, for some
{\color{blue} (even polynomial)} GADTs,
%that are not hereditarily polynomial nested types,
no such isomorphism holds.

GADTs are traditionally represented via Church encodings. In this
view, GADTs are completely determined by their syntax. For example,
the GADT $\mathsf{G}$ defined by
\[\begin{array}{l}
\mathsf{data\,G : * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
\end{array}\]
{\color{blue} Use $\top$ instead of $1$ throughout?}

@@

A GADT programmer is likely to use GADTs precisely because they
exhibit different behaviors at different types, and thus to consider
the GADT to be completely specified by its syntax. When used this way,
the shape of a GADTs is actually {\em determined} by the data it
contains. As a result, it is not expected to support a $\mathsf{map}$
function --- i.e., to be an instance of Haskell's $\mathsf{Functor}$
class --- like other data types. For example, the definition for
$\mathsf{G}$ above specifies data only at instance $\mathsf{G\,1}$.
Since there is a function from $\mathsf{1}$ to any other type
$\mathsf{t}$, and yet $\mathsf{G\,t}$ is uninhabited at any instance
other than $\mathsf{1}$, $\mathsf{G}$ cannot support a $\mathsf{map}$
function. When it is intended to capture this kind of non-uniform
behavior, a GADT's behavior does not generalize that of ADTs as
advertised, and in some sense doesn't even behave like a traditional
data type.  When used this way, a GADT cannot be seen as the least
fixpoint of a functor, and must instead be represented via its Church
encoding.

@@

Consider, again, the GADT $\mathsf{Seq}$ from the start of
Section~\ref{sec:gadts}. Because its type argument varies in the
instances of $\mathsf{Seq}$ appearing in the types of its data
constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be characterized by
any first-order fixpoint equation. It can, however, be viewed as a
solution to the higher-order fixpoint equation
\[H\,f\,a = a \,+\,Lan_{\lambda a b. a \times b}\, \lambda
a b. f a \times f b\] where $\Lan_K \,G$ is the left Kan extension of
$G$ along $K$.  The left Kan extension $\Lan_K \,G : {\cal E} \to
{\cal D}$ of the functor $G : {\cal C} \to {\cal D}$ along the functor
$K : {\cal C} \to {\cal E}$ is the best functorial approximation to
$G$ that factors through $K$. Here, ``best functorial approximation''
means that $\Lan_K \,G$ is the smallest functor that both extends the
image of $K$ to $\cal D$ and coincides with $G$ on $\cal C$, in the
sense that, for any other such functor $H$, there is a morphism of
functors (i.e., a natural transformation) from $\Lan_K \,G$ to
$H$. This is captured formally by the following
definition~\cite{mac71}:
\begin{definition}\label{def:lke}
If $G : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
functors, then the {\em left Kan extension of $G$ along $K$} is a
functor $\mathit{Lan}_K\,G : {\cal E} \to {\cal D}$ together with a
natural transformation $\eta : G \to (\mathit{Lan}_K\,G) \circ K$
such that, for every functor $H : {\cal E} \to {\cal D}$ and natural
transformation $\gamma : G \to H \circ K$, there exists a unique
natural transformation $\delta : \mathit{Lan}_K\,G \to H$ such that $(\mu
K) \circ \eta = \gamma$. This is depicted in the diagram
\[\begin{tikzcd}[row sep = huge]
{\cal C}
\ar[rr, "{G}"{name=Fa, above}, ""{name=F, below}]
\ar[rd, "{K}"']
&& {\cal D} \\
& {\cal E}
\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
\ar[ur, bend left, "{\mathit{Lan}_{K}\,G}"{name=Lan, description}]
\ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\delta}"']
\end{tikzcd}\]
\end{definition}

To represent GADTs as primitive data types a calculus must support a
primitive fixpoint constructor $\mathsf{\mu}$ that is the syntactic
reflection of the higher-order fixpoint operator on the category of
functors on the cagegory interpretating types, as well as a primitive
type constructor $\mathsf{Lan}$ such that $\mathsf{Lan_k\,g}$ is the
syntactic reflection of the left Kan extension $\Lan_K \,G$ of the
functor $G$ interpreting $\mathsf{g}$ along the functor $K$
interpreting $\mathsf{k}$. {\color{blue} This is the least commitment
  thing primitive representations can require. Fixpoints were elided
  in ADT case. Why? Tie together better.}  The type constructor
$\mathsf{Lan}$ thus computes the smallest data type that is generated
by a GADT's {\color{blue} GADT or constructor?} syntax and also
supports a $\mathsf{map}$ function. That is, $\mathsf{Lan}$ computes
the functorial completion of the data type determined by a GADT's
syntax, filling it out with data elements that are derivable via
repeated applications of $\mathsf{map}$ starting from those elements
specified by its syntax. In the primitive view, GADTs are, in general,
underspecified by their syntax.

Just as for ADTs, we can represent GADTs as either Church encodings or
primitive data types. But whereas the two representations are always
isomorphic for ADTs, they not, in general, for GADTs. We will see this
in the next section.

@@

Intuitively it seems the two representations might be different. One
is functorial, one is not. Non-uniform nature of GADTs renders build,
therefore short cut, for GADTs impossible. Can't even formulate! Of
course, there could be another way to prove equivalence.  But, in
fact, CEs and primitive representations are not equivalent for
arbitrary GADTs, as the following example shows.

\begin{example}
Use our favorite GADT to establish non-equivalence of the two reps
\end{example}

Forced to choose, a programmer would likely find the idea that a GADT
contains data that are not specified by its syntax disturbing. What,
they might ask, should a data type contain other than data that are
constructed using its data constuctors? Why should a GADT contain
``secret'' elements that are only attainable by mapping? On the other
hand, a semanticist would likely find the primitive representation of
GADTs more reasonable. After all, they would likely find the
nonfunctorial nature of the Church encoding unnerving at best. After
all, they would argue, the data in a GADT shouldn't disappear just
because you map a function over it! The fact that this happens for
Church encodings of GADTs --- i.e., for GADTs as specified by their
syntax --- highlights how GADTs {\em do not} generalize the essential
container-ish nature of ADTs at all. A semanticist might therefore
conclude that GADTs are seriously misnamed. They might even go on to
wonder: in what sense are GADTs data types at all?

@@

\section{Parametricity in the Presence of GADTs}

@@

1. Parametricity requires a fibred semantics. This doesn't happen for
primitive rep.

2. What are the parametricity properties for GADTs in each setting?
(Naturality only with primitives, inhabitation etc. with CEs)

@@

We do minimal thing to fix functoriality, then investigate
parametricity properties for both interpretations. We are forced to
choose.

@@

GADTs Force a Representation Choice. It has consequences for how one
reasons about them.

>>>>

There are two standard ways to represent data types when studying
modern functional languages: via their Church encodings in a (possibly
higher-kinded) extension of System F~\cite{gir72}, the calculus at the
core of all such languages, and by augmenting System F
%a Hindley-Milner style calculus
with primitives for constructing them directly as fixpoints.
%{\color{blue} Something about when each of these representations is
%  used. The former perhaps more closely represents how data types are
%  represented in implementations, but the latter are how they have
%  heretofore primarily been treated theoretically.}
In models that validate the program optimization known as {\em short
  cut fusion}~\cite{glp93} the two representations are, in fact,
equivalent, in the sense that they both have the same interpretation
in the model. Models in which the Church encodings and primitive
representations of all {\color{blue} (polynomial)} algebraic data
types (ADTs) are equivalent include the operationally-based model
of~\cite{pit98,pit00} and the categorical models
of~\cite{joh02,joh03}.
%\footnote{{\color{blue} Pitts proves the isomorphism only for
%    polynomial ADTs, whereas Johann {\em et al.}  prove it for any
%    data type that is the fixpoint of a first-order functor.}}
This equivalence was recently extended to the class of what might be
called ``hereditarily polynomial'' {\color{blue} to include truly
  nested such types} nested types via the categorical model
of~\cite{jgj21}. But, as shown in Section~\ref{sec:par} below, for
some {\color{blue} (even polynomial)} GADTs,
%that are not hereditarily polynomial nested types,
no such isomorphism holds.

{\color{red} Maybe use a less degenerate example.}  GADTs are
traditionally represented via Church encodings. In this view, GADTs
are completely determined by their syntax. For example, the GADT
$\mathsf{G}$ defined by
\[\begin{array}{l}
\mathsf{data\,G : * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
\end{array}\]
{\color{blue} Use $\top$ instead of $1$ throughout?}

@@

{\color{red} This will probably be more convincing if we use, e.g.,
  an expression GADT from Stephanie above.}
A GADT programmer is likely to use GADTs precisely because they
exhibit different behaviors at different types, and thus to consider
the GADT to be completely specified by its syntax. When used this way,
the shape of a GADTs is actually {\em determined} by the data it
contains. As a result, it is not expected to support a $\mathsf{map}$
function --- i.e., to be an instance of Haskell's $\mathsf{Functor}$
class --- like other data types. For example, the definition for
$\mathsf{G}$ above specifies data only at instance $\mathsf{G\,1}$.
Since there is a function from $\mathsf{1}$ to any other type
$\mathsf{t}$, and yet $\mathsf{G\,t}$ is uninhabited at any instance
other than $\mathsf{1}$, $\mathsf{G}$ cannot support a $\mathsf{map}$
function. When it is intended to capture this kind of non-uniform
behavior, a GADT's behavior does not generalize that of ADTs as
advertised, and in some sense doesn't even behave like a traditional
data type.  When used this way, a GADT cannot be seen as the least
fixpoint of a functor, and must instead be represented via its Church
encoding.

@@

Consider, again, the GADT $\mathsf{Seq}$ from the start of
Section~\ref{sec:gadts}. Because its type argument varies in the
instances of $\mathsf{Seq}$ appearing in the types of its data
constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be characterized by
any first-order fixpoint equation. It can, however, be viewed as a
solution to the higher-order fixpoint equation
\[H\,f\,a = a \,+\,Lan_{\lambda a b. a \times b}\, \lambda
a b. f a \times f b\] where $Lan_K \,G$ is the left Kan extension of
$G$ along $K$.  The left Kan extension $Lan_K \,G : {\cal E} \to
{\cal D}$ of the functor $G : {\cal C} \to {\cal D}$ along the functor
$K : {\cal C} \to {\cal E}$ is the best functorial approximation to
$G$ that factors through $K$. Here, ``best functorial approximation''
means that $Lan_K \,G$ is the smallest functor that both extends the
image of $K$ to $\cal D$ and coincides with $G$ on $\cal C$, in the
sense that, for any other such functor $H$, there is a morphism of
functors (i.e., a natural transformation) from $Lan_K \,G$ to
$H$. This is captured formally by the following
definition~\cite{mac71}:
\begin{definition}\label{def:lke}
If $G : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
functors, then the {\em left Kan extension of $G$ along $K$} is a
functor $\mathit{Lan}_K\,G : {\cal E} \to {\cal D}$ together with a
natural transformation $\eta : G \to (\mathit{Lan}_K\,G) \circ K$
such that, for every functor $H : {\cal E} \to {\cal D}$ and natural
transformation $\gamma : G \to H \circ K$, there exists a unique
natural transformation $\delta : \mathit{Lan}_K\,G \to H$ such that $(\mu
K) \circ \eta = \gamma$. This is depicted in the diagram
\[\begin{tikzcd}[row sep = huge]
{\cal C}
\ar[rr, "{G}"{name=Fa, above}, ""{name=F, below}]
\ar[rd, "{K}"']
&& {\cal D} \\
& {\cal E}
\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
\ar[ur, bend left, "{\mathit{Lan}_{K}\,G}"{name=Lan, description}]
\ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\delta}"']
\end{tikzcd}\]
\end{definition}

To represent GADTs as primitive data types a calculus must support a
primitive fixpoint constructor $\mathsf{\mu}$ that is the syntactic
reflection of the higher-order fixpoint operator on the category of
functors on the category interpretating types, as well as a primitive
type constructor $\mathsf{Lan}$ such that $\mathsf{Lan_k\,g}$ is the
syntactic reflection of the left Kan extension $\Lan_K \,G$ of the
functor $G$ interpreting $\mathsf{g}$ along the functor $K$
interpreting $\mathsf{k}$. Using the $\mu$ and $\Lan$ type
constructors we can represent, e.g., the GADT $\mathsf{Seq}$ as
\[\mathsf{Seq\,a} = \mathsf{\mu \phi.\, a + Lan_{\lambda c d. c \times d}
  \lambda c d. \phi c \times \phi d}\]

{\color{blue} This is the least commitment
  thing primitive representations can require. Fixpoints were elided
  in ADT case. Why? Tie together better.}  The type constructor
$\mathsf{Lan}$ thus computes the smallest data type that is generated
by a GADT's {\color{blue} GADT or constructor?} syntax and also
supports a $\mathsf{map}$ function. That is, $\mathsf{Lan}$ computes
the functorial completion of the data type determined by a GADT's
syntax, filling it out with data elements that are derivable via
repeated applications of $\mathsf{map}$ starting from those elements
specified by its syntax. In the primitive view, GADTs are, in general,
underspecified by their syntax.

Just as for ADTs, we can represent GADTs as either Church encodings or
primitive data types. But whereas the two representations are always
isomorphic for ADTs, they are not, in general, for GADTs. We will see
this in the next section.

@@

Intuitively it seems the two representations might be different. One
is functorial, one is not. It is possible they could always have the
same data elements even if not the same morphisms. But as we see in
Example~\ref{ex:g1}, this turns out not to be the case.

Non-uniform nature of GADTs renders build, therefore short cut, for
GADTs impossible. Can't even formulate! Of course, there could be
another way to prove equivalence.  But, in fact, CEs and primitive
representations are not equivalent for arbitrary GADTs, as the
following example shows.

\begin{example}\label{ex:g1}
Use our favorite GADT to establish non-equivalence of the two reps
\end{example}

Forced to choose, a programmer would likely find the idea that a GADT
contains data that are not specified by its syntax more than a little
disturbing. What, they might ask, should a data type contain other
than data that are constructed using its data constuctors? Why should
a GADT contain ``hidden'' elements that are only accessible by
mapping? On the other hand, a semanticist would likely find the
primitive representation of GADTs more reasonable. Indeed, they would
likely find the nonfunctorial nature of the Church encoding unnerving
at best. After all, they would likely argue, the data in a GADT
shouldn't disappear just because you map a function over it! The fact
that this happens when GADTs are represented by their Church encodings
actually highlights how GADTs {\em do not} generalize the essential,
container-ish nature of ADTs at all. A semanticist might therefore
conclude that GADTs are seriously misnamed. They might even go on to
wonder: in what sense are GADTs data types at all?

@@

\section{Parametricity in the Presence of GADTs}\label{sec:par}

@@

1. Parametricity requires a fibred semantics. This doesn't happen for
primitive rep so can't have a par model with this rep. Can still get
naturality properties, however.

2. For CEs can get par model and par properties. But what do these say
for GADTs? Give examples?

@@

@@
